---
title: "(Very) Preliminary Study of the Longevity of Docker Images from Research Artifacts"
author: Quentin Guilloteau
geometry: margin=20mm
numbersections: true
bibliography: references.bib
---


```{r, include=F}
#
library(tidyverse)
library(slider)
```

```{r}
df <- read_csv("./outputs/aggregated/europar24/pkgs.csv", col_names = c("pkg", "version", "tool", "artifact", "date")) %>%
    mutate(date = date(as_datetime(date))) %>%
    drop_na()
df
```


# Introduction


- Reproducibility crisis

- Artifact evaluation in conferences

- Software environment is a problem/difficul aspect

- conferences recommand to use containers to share the software environmnent

- but container recipes are based on non-reproducibile tools (package managers and bad practices) @guilloteau:hal-04562691

- trade-off between storing the result of the image builts by the authors and the recipe


This paper proposes a framework to evaluate the variations in resulting software environment produced by `Dockerfile`s from research artifacts.
We show initial results from five `Dockerfile`s built every month for six months.

# Methodology

In this paper, we build Docker images coming from `r length(unique(df$artifact))` artifacts of the EuroPar'24 conference.
We built each image on `r length(unique(df$date))` dates roughtly one month apart (`r unique(df$date)`).


We aim to answer the following questions:

1. How does the packages evolve through time when captured in a `Dockerfile`?

2. Which packaging tool is the most victim to variation in its resulting software environment?

3. Which are the most volatile packages?


The goal is not to dunk on authors or conferences, but simply to exhibit the potential issues with `Dockerfile`s.

# Framework

- `ecg`

- architecture (nickel, what is supported)

- process (submit new artifacts)

- end goal of the project

# Results

This section presents the information we would like to extract from the data gathered.


```{r}
df %>%
    group_by(pkg, tool, artifact) %>%
    summarize(n = n_distinct(version)) %>%
    arrange(desc(n)) %>%
    ggplot(aes(x = n, color = tool)) +
    stat_ecdf() +
    xlim(1, NA) +
    xlab("Number of different versions") +
    facet_wrap(~artifact) +
    theme_bw()
```

```{r}
first_dates <- df %>%
    group_by(artifact) %>%
    summarize(first_date = min(date))

first_date_versions <- df %>%
    left_join(first_dates) %>%
    filter(first_date == date) %>%
    rename(first_version = version) %>%
    select(first_date, pkg, first_version, artifact)

df %>%
    left_join(first_date_versions) %>%
    mutate(is_same_as_first_date = version == first_version) %>%
    select(date, pkg, artifact, tool, is_same_as_first_date) %>%
    group_by(artifact, date) %>%
    mutate(total_pkgs = n()) %>%
    group_by(artifact, date, is_same_as_first_date, total_pkgs) %>%
    summarize(total = n()) %>%
    ggplot(aes(x = factor(date), y = 100 * total / total_pkgs, fill = is_same_as_first_date)) +
    geom_col() +
    # geom_text(data = .%>% filter(is_same_as_first_date), aes(y = (100 * total/total_pkgs) - 1.5, label = paste(round(100*total/total_pkgs, digits=1), "%",sep="")), angle=0, size=1.6) +
    geom_text(data = .%>% filter(is_same_as_first_date), aes(y = (100 * total/total_pkgs) - 2, label = total), angle=0, size=2.0) +
    geom_text(data = .%>% filter(is_same_as_first_date), aes(y = (100 * total/total_pkgs) + 2, label = total_pkgs - total), angle=0, size=2.0) +
    #geom_hline(yintercept = 85, linetype="dashed") +
    xlab("Dates of builds (YYYY-MM-DD)") + 
    ylab("Percentage of the packages [%]") +
    scale_fill_discrete("Are the versions of the packages the same as the initial build?", breaks = c(TRUE, FALSE), labels = c("Yes", "No")) +
    facet_wrap(~artifact, ncol = 5) +
    ggtitle("Evolution of the packages versions through time") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
```
```{r}
ndf <- df %>% 
  group_by(artifact, tool, pkg) %>%
  arrange(date) %>%
  nest()

xx <- ndf %>%
  rowwise() %>%
  mutate(data2 = list(data %>% arrange(date) %>% mutate(rank = .$date[match(version, .$version)]))) %>%
  select(-data) %>%
  unnest_wider(data2) %>%
  unnest_longer(c("version", "date", "rank"))

dates <- xx %>% pull(date) %>% unique()
xx %>%
  group_by(artifact, date) %>%
    mutate(total_pkgs = n()) %>%
    group_by(artifact, date, rank, total_pkgs) %>%
    summarize(total = n()) %>%
    mutate(rank = factor(rank)) %>%
    mutate(rank = factor(rank, levels=rev(levels(rank)))) %>%
    ggplot(aes(x = factor(date), y = 100 * total / total_pkgs, fill = rank)) +
    geom_col() +
    # geom_text(data = .%>% filter(is_same_as_first_date), aes(y = (100 * total/total_pkgs) - 1.5, label = paste(round(100*total/total_pkgs, digits=1), "%",sep="")), angle=0, size=1.6) +
    #geom_hline(yintercept = 85, linetype="dashed") +
    geom_text(data = . %>% filter(as.character(rank) == dates[1]), aes(y = (100 * total/total_pkgs) - 5, label = total), angle=0, size=2.0) +
    geom_text(data = . %>% filter(as.character(rank) == dates[1]), aes(y = (100 * total/total_pkgs) - 10, label = paste("(",round(100*total/total_pkgs, digits = 1), "%)",sep="")), angle=0, size=1.5) +
    #xlab("Dates of builds (YYYY-MM-DD)") + 
    ylab("Packages in environment [%]") +
    facet_wrap(~artifact, ncol = 5) +
    ggtitle("Evolution of the packages versions through time") +
    scale_fill_grey("Month when the package version was introduced in the environment", labels=rev(append(c("Initial"), seq(length(dates)-1))))+
    scale_x_discrete("Months after initial build", labels=append(c("Initial"), seq(length(dates)-1))) +
    #scale_fill_grey() +
    theme_bw() +
    theme(legend.position = "bottom", strip.background = element_blank())
    #theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom", strip.background = element_blank())

ggsave("rep24/fig.png", width=8, height=3.5)
```

```{r}
ps <- df %>%
    left_join(first_date_versions) %>%
    mutate(is_same_as_first_date = version == first_version) %>%
    select(date, pkg, artifact, tool, is_same_as_first_date) %>%
    group_by(artifact, date) %>%
    mutate(total_pkgs = n()) %>%
    group_by(artifact, date, is_same_as_first_date, total_pkgs) %>%
    summarize(total = n()) %>%
    mutate(r = total / total_pkgs) %>%
    filter(is_same_as_first_date) %>%
    select(-total_pkgs, -total, -is_same_as_first_date) %>%
    ungroup() %>%
    group_by(artifact) %>%
    mutate(p = r / lag(r)) %>%
    drop_na() %>%
    select(artifact, p)

qqnorm(ps$p, main='Normal')
qqline(ps$p)
ks.test(ps$p, 'pnorm')
summary(ps$p)
```

```{r}
df %>%
    left_join(first_date_versions) %>%
    mutate(is_same_as_first_date = version == first_version) %>%
    select(date, pkg, artifact, tool, is_same_as_first_date) %>%
    group_by(artifact, tool, date) %>%
    mutate(total_pkgs = n()) %>%
    group_by(artifact, date, tool, is_same_as_first_date, total_pkgs) %>%
    summarize(total = n()) %>%
    ggplot(aes(x = factor(date), y = 100 * total / total_pkgs, fill = is_same_as_first_date)) +
    #ggplot(aes(x = factor(date), y = total, fill = is_same_as_first_date)) +
    geom_col() +
    #geom_text(data = .%>% filter(is_same_as_first_date), aes(y = 100*(total/total_pkgs) - 6, label = total), angle=0, size=2.0) +
    #geom_text(data = .%>% filter(is_same_as_first_date), aes(y = 100*(total/total_pkgs) + 6, label = total_pkgs - total), angle=0, size=2.0) +
    xlab("Dates of builds (YYYY-MM-DD)") + 
    ylab("Percentage of the packages") +
    scale_fill_discrete("Are the versions of the packages the same as the initial build?", breaks = c(TRUE, FALSE), labels = c("Yes", "No")) +
    facet_grid(tool~artifact, scales="free_y") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
```

```{r}
# xx %>%
#     #left_join(first_date_versions) %>%
#     #mutate(is_same_as_first_date = version == first_version) %>%
#     #select(date, pkg, artifact, tool, is_same_as_first_date) %>%
#     group_by(tool, date, rank) %>%
#     mutate(total_pkgs = n()) %>%
#     group_by(artifact, date, tool, is_same_as_first_date, total_pkgs) %>%
#     summarize(total = n()) %>%
#     ggplot(aes(x = factor(date), y = 100 * total / total_pkgs, fill = is_same_as_first_date)) +
#     geom_col() +
#     ylab("Percentage of the packages") +
#     facet_wrap(~tool) +
#     scale_fill_grey("Month when the package version was introduced in the environment", labels=rev(append(c("Initial"), seq(length(dates)-1))))+
#     scale_x_discrete("Months after initial build", labels=append(c("Initial"), seq(length(dates)-1))) +
#     theme_bw() +
#     theme(legend.position = "bottom")


xx %>%
  group_by(tool, date) %>%
    mutate(total_pkgs = n()) %>%
    group_by(tool, date, rank, total_pkgs) %>%
    summarize(total = n()) %>%
    mutate(rank = factor(rank)) %>%
    mutate(rank = factor(rank, levels=rev(levels(rank)))) %>%
    ggplot(aes(x = factor(date), y = 100 * total / total_pkgs, fill = rank)) +
    geom_col() +
    # geom_text(data = .%>% filter(is_same_as_first_date), aes(y = (100 * total/total_pkgs) - 1.5, label = paste(round(100*total/total_pkgs, digits=1), "%",sep="")), angle=0, size=1.6) +
    #geom_hline(yintercept = 85, linetype="dashed") +
    geom_text(data = . %>% filter(as.character(rank) == dates[1]), aes(y = (100 * total/total_pkgs) - 8, label = total), angle=0, size=2.5) +
    geom_text(data = . %>% filter(as.character(rank) == dates[1]), aes(y = (100 * total/total_pkgs) - 24, label = paste("(",round(100*total/total_pkgs, digits = 1), "%)",sep="")), angle=0, size=2) +
    #xlab("Dates of builds (YYYY-MM-DD)") + 
    ylab("Packages in environment [%]") +
    facet_wrap(~tool, ncol = 2) +
    ggtitle("Evolution of the packages versions through time") +
    scale_fill_grey("Month when the package version was introduced in the environment", labels=rev(append(c("Initial"), seq(length(dates)-1))))+
    scale_x_discrete("Months after initial build", labels=append(c("Initial"), seq(length(dates)-1))) +
    #scale_fill_grey() +
    theme_bw() +
    theme(legend.position = "bottom", strip.background = element_blank())
    #theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom", strip.background = element_blank())

ggsave("rep24/fig2.pdf", width=5, height=3.5)
```
## Most volatile packages 


```{r}
df %>%
    group_by(pkg, artifact, tool) %>%
    summarize(different_versions = n_distinct(version)) %>%
    ungroup() %>%
    slice_max(different_versions, n=10) %>%
    knitr::kable()

df %>%
    group_by(pkg, artifact, tool) %>%
    summarize(different_versions = n_distinct(version)) %>%
    ggplot(aes(x = different_versions, color = tool)) +
    stat_ecdf() + 
  xlim(0, 6) +
  theme_bw()
```

```{r}
pname = "linux-libc-dev:amd64"

libc_versions <- df %>%
    filter(pkg == pname) %>%
    pull(version) %>%
    unique()
```


```{r}
df_libc <- read_delim("./ubuntu_linux-libc-dev_changelogs/all_versions.csv", delim=";;", col_names=c("version", "date")) %>%
    mutate(date = as_date(date))

build_dates <- tibble(bdate = unique(df$date))

df_libc %>%
    filter(date >= min(df$date) - months(6)) %>%
    print() %>%
    mutate(was_built = if_else(version %in% libc_versions, "built", "not built")) %>%
    ggplot() +
    geom_point(aes(x = date, y = 0, color = was_built)) +
    geom_vline(data = build_dates, aes(xintercept = bdate), linetype = "dashed", color = "gray") +
    theme_bw()
#libc_versions
unique(df_libc %>% filter(date >= min(df$date) - months(6)) %>% pull(version))
```

# Threats to validity

Very small sample, only Docker.

# Conclusion

`Dockerfile`s are not solutions to capture on the long term the software environment.
There is a trade-off between storing the Docker images built by the authors and their recipe.
Use Nix/Guix + Software Heritage.













